%!TEX encoding = UTF-8 Unicode
%preamble
\documentclass{article}
\usepackage{caption}
\usepackage{titlesec}
\usepackage[top=1in, left=1in]{geometry}
\usepackage{fancyhdr}
%fancy header
    \pagestyle{fancy}
    \lhead{COSI 216 A}
    \chead{\textbf{`Support' Group Contract}}
    \rhead{\today}
    \lfoot{}
    \cfoot{}
    \rfoot{}

%Signature Lines
\newcommand{\sigline}[2]{\vspace{2em} #1 \hfill \line(1,0){#2}}

%document
\begin{document}

\paragraph{Group Members} % (fold)
\label{par:group_members}
\begin{center}
    This document represents an agreement between the following parties:\\
    \emph{Cory Massaro}\\
    \emph{Cynthia Goodman}\\
    \emph{Jasper Phillips}\\
    \& \emph{Zachary Yocum}    
\end{center}
% paragraph group_members (end)

\paragraph{Expectations} % (fold)
\label{par:expectations}
The expectation is that each member will attend weekly group meetings and meet all assigned deadlines. The schedule for such meetings is yet to be determined.
% paragraph expectations (end)

\section{Skills} % (fold)
\label{sec:skills}

The relevant skills that each member believes will aid in their contributions are as follows:

\begin{description}
    \item[Cory:] Programming in C, Java, and Python as well as Inter-Annotator Agreement (IAA) $\kappa$ score calculations.
    \item[Cynthia:] Programming in Java and Python. Experience with morpho-syntactic annotation of American Sign Language.
    \item [Jasper:] Programming in Python and document layout compilation with \LaTeX.
    \item[Zach:] Programming in Java and Python. Experience with Multi-document Annotation Environment (MAE). Experience with document layout compilation with \LaTeX.
\end{description}
% section skills (end)

\section{Roles} % (fold)
\label{sec:roles}

Roles within the group will be designated as follows:

\begin{description}
    \item [Notetaker:] Jasper
    \item [Editors:] Cory, Cynthia
    \item [Compiler:] Zach
\end{description}
% section roles (end)

\paragraph{Agreement} % (fold)
\label{par:agreement}
The undersigned hereby agree, to the best of their respective abilities and skills as outlined in Section \ref{sec:skills}, to contribute equally to the project according to their roles as designated in Section \ref{sec:roles}:\footnote{\emph{Disclaimer}: This is a preliminary group contract. As our understanding of the project becomes more clear we reserve the option to revise and resubmit a modified contract to reflect that understanding.}
% paragraph agreement (end)
\vspace{3em}\\
\sigline{Cory Massaro}{300}\\
\sigline{Cynthia Goodman}{300}\\
\sigline{Jasper Phillips}{300}\\
\sigline{Zachary Yocum}{300}\\
%\vspace{2em} Cory Masaro \hfill \line(1,0){300}\\
%\vspace{2em} Cynthia Goodman \hfill \line(1,0){300}\\
%\vspace{2em} Jasper Phillips \hfill \line(1,0){300}\\
%\vspace{2em} Zachary Yocum \hfill \line(1,0){300}
\end{document}
